\chapter{Solving a quadratic equation}
\begin{quotation}\textit{
Enthusiastic educators can set tasks in introductory computing couses
that give their students a taste of program creation but are not
too compilicated or difficult. Until you demand that the answers
really appoach perfection. The example given here is just to give
a taste of this!
}\end{quotation}
There are introductions to programming that give as one of their
earliest examples the challange of creating an application that
reads in three numbers, $a$, $b$ and $c$ and then prints out the
two solutions to the equation $a x^2 + b x + c = 0$. The clear
expectation is that this will be done using the well-known formula
$\frac{-b \pm \sqrt{b^2 - 4 a c)}}{2 a}$. When done by just writing out
use of that formula the task is indeed easy enough to be in an
introductory section about use of computers.

However if one looks at the task more carefully and start to insist on
getting as correct a result as possible for all legitimate inputs things
become rather different. I will suppose that since this is a task
set for beginners it is to be solved using the default way in which
your computer handles numbers. In almost all cases this will follow
an international standard called IEE~754 abd the program that is written
will use a representation they call ``binary64'' which is commonly referrred
to as ``double precision''. The task in hand here highlights several
ways in which the computer arithmetic that is thereby provided can
do things that a naive user might not havr thought about. So let's
take these in turn and see how the equation solver will need to
be made more elaborate to avoid them hurting.

The (IEEE) floating point representation in a computer can not handle
arbitrarily large numbers. Specifically it runs out of steam when values
exceed around $1.8 \times 10^{308}$. Beyond that it stores values
that represent ``infinity''. Now of course sensible people will
not tend to work with problems that lead to numbers so absurdly large,
and so it is common not to think much about it! However if one seeks
perfection then the quadratic solver should deliver accurate results for
any inputs where the results are sensible, and it should report
failure in exactly the cases where the inputs do not lead to answers that
can be represented in the floating point format that the computer uses.

So now consider the case when some user provides the input
$a=1, b=1, c=-1$. Substituting these into the formula leads to
a pair of very sensible roots with values about $0.618034$ and
$-1.618034$. Everything seems good. But now a rather less helpful
user enthusiastically offers $a=b=2 \times 10^{154}$ and
$c=-2 \times 10^{154}$. This fairly obviously has exactly the same
two roots. However the with this and with various fairly closely related
cases involving huge numbers it is possible to arrange that in the
computation of $b^2 - 4 a c$ that either $b^2$ overflow or $4 a c$
overflow or both, or that neither overflow but their difference does.
It is furthemore possible for one or both of the terms to exceed the
$1.8 \times 10^{308}$ maximum and hence overflow even though when they
are combined the result is in range. This is all a mess! However
this particular case can be tidied up by starting the calculation by
dividing all of $a$, $b$ and $c$ by some large value so as to reduce them
to sensible size numbers. At which point an additional issue comes into
play. If you divide a floating point number by a general scale factor
doing so can introduce rounding errors. Consider for instance the
calculation of just $1/3$ and the computer will produce something
in the style of $0.3333333333333333$\footnote{Since it is working in binary
internally it will actually be more like $0.01010101\ldots$} which is
not precisely the same since it terminates after a finite number of
digits. Continuing the calculation with this corrupted value will
naturally lead to a (small) error in the final result. Happily with
IEEE floating point it is legitimate to divide by any power of $2$ and
in that case no precision will be lost. So it will be necessary to
identify a power of $2$ that is about the right size so that it can be used
to rescale the input to avoid overflow.

The situation with very very small numbers is in fact even more curious
although the way to sort things out is basically the same. The smallest
floating point value that can exist (with smaller values being flushed
to zero) is about $4.94\times 10^{-324}$\footnote{$2^{-1917}$} but once
values get lower than $2.23 \times 19^{-308}$ the representation starts to
work with them an lower than normal precision. So to preserve as
much accuracy as possible the scaling has to keep things away from not the
point where underflow maps values to zero, but from some rather
earlier-encountered threshold.

For now and to prevent things geting quite out of hand the issues of
rounding errors that might arise when multiplying $b$ by itself and so
on up to and including those that could be introduced by the square root
calculation are going to be ignored, but anybody who really wants to
push for perfection regardless of the cost in difficulty can explore
those paths. Also it would be proper to review cases where the true
results are going to be very close to or beyond the overflow or underflow
points so that good answers are produced in every case where that is possible.
Part of that might involve finding a scale factor $\sigma$ and replacing $a$
with $\sigma a$ and $c$ with $c / \sigma$\footnote{with $\sigma$ a power of
2 to avoid introducing corruption through rounding}. Then a final result
can be generated by multiplying or dividing the solutions to the
scaled version by $\sigma$, and any overflow or underflow will then
be captured in and limited to that final re-scaling operation.

However there is a further and distinct way in which the naive use of the
standard formula can generate seriously inaccurate results even after
overflow is avoided and regardless of minor rounding errors.
Supose that the value of $-b$ and $\sqrt{b^2 - 4 a c}$ are rather similar
in absolute value. This easily gappens when $b$ is rather larger than
either $a$ or $c$. Then one of their sum and their difference will
add two similar values and give a good result, while the other can lead
to massive loss of precision as leading digits match and cancel out. To
illustrate this consider the use of 8-digit decimal floating point on
a pocket calculation and look at the equation $x^2 - 4003*x -1 = 0$.
Here $-b$ is obviously just $4003.0000$ when written so that the 8
digits of precison are made explicit. A careful  calculation of
$\sqrt{b^2 - 4 a c}$ gives its value as $4003.000499625249859\ldots$ and
to 8 digits that is $4003.0005$. When the computer subtracts these it has
no access to ant of the lower digits so the only result it can offer
is $-0.0005$ while the ideal result would have been $-0.00049962524$, So
although the arithmetic had been being performed keeping 8 significant
digits thoughout the cancellation of leading digits in the subtration
means that our final result has at best only 4 of its leading digits
correct. Using a number larger than $4003$ would make this precision loss
worse to an extent that even with full 64-bit IEEE arithmetic results
can be unsatisfactory.


While we can all accept that the example shown here might have been chosen
with awkward numbers deliberately picked to give this severe cancellation
of loading digits, there is no fundamental reason why such cases might not
arise in real life and in fact it can occur so easily that it should be
considered a common risk.

Happily (for those who like to deliver accurate anwers) or painfully (for
those who see this adding yet an extra layer of complication and
difficulty to the task) it is possible to avoid this calamity, because
it is known that the product of the two roots of a quadratic will
always be equal to $c/a$ and bacause whenever calculating one of the two roots
does a subtraction that can lead to leading digit cancellation the other
will be found by doing an addition that gives full precision results. So
in effect one should calculate the more delicate root as
$2 c/(-b - \sqrt{b^2 - 4 a c}))$ which will now be very respectable.
OF course it will be necessary to judge which of the plus or minus cases
is the one to use and which the one to avoid.

So overall the program that solves a simple quadratic and that does not
even worry about cases where there is no (real) solution is dramatically
messier and calla for much more understanding that those novice
programmers will have been ready to deploy. But if you are writing a library
or application that is to be used by others you have a responsibility to
cope with all cases, not just the ones you think about first!


 